---
title: "Machine (Statistical) Learning and potential application to chronic disease conditions"
subtitle: ""
author: "Paola Berchialla <br /> <br /> <br /> Department of Clinical and Biological Sciences <br />  University of Torino"
date: "2018-10-09"
output:
  xaringan::moon_reader:
    css: [default, my_theme.css, default-fonts]
    lib_dir: libs
    mathjax: null
    chakra: libs/remark-latest.min.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Machine Learning in healthcare 
```{r, echo = FALSE, message = FALSE, warning=FALSE, fig.align='center', fig.height=5}
library(magrittr)
library(dplyr)
library(ggplot2)
load('ml_pubmed.RData')


fit <- lm(log(Publications)~Year, data = ml_pubmed)
new_df = data.frame(Year = c(2000:2018,seq(2018,2021, length.out = 100)))
predicted_df <- data.frame(Year = new_df$Year, 
                           Publications = exp(predict(fit,new_df )),
                           Timepoint = c(rep('Current', length(c(2000:2018))),
                                         rep('Projected', 100)))

ggplot(ml_pubmed, aes(x = Year, y = Publications))+
  geom_line( data = predicted_df,
            aes(x = Year, y = Publications, col = Timepoint), lwd = 1.5)+
  geom_point(size = 2)+
  geom_vline(xintercept = 2018, lty = 'dashed', lwd = 1)+
  geom_text(x = 2010, y = 6000, label = "R^2 == 0.9965", parse = T)+
  geom_text(x = 2010, y = 5000, label = "log(Publications) %~% Year", parse = T)
```

- Growth of publications in PubMed with term _machine learning_
- Exponentiated regression of log number of publications on year is used to predict the future trend 

---
# What is Machine Learning

- Machine learning teaches computers to do __learn__ from experience
    - Machine learning algorithms use computation methods to _learn_ information directly from data without relying on a predetermined equation to model

- The algorithms adaptively improve their performance as the number of data samples available for learning increases


---
# How Machine Learning works

__Different types of Learning__

- Supervised Learning
  - output is known and it is used during the training process
- Unsupervised Learning
  - output is not provided during the training
- Other type of Learning
  - Semi-Supervised Learning
  - Reinforcement Learning
.pull-left[]
.pull.right[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='right', out.width="60%"}
library(knitr)
img_path <- "images/Learning.png"
include_graphics(img_path)
```
]

---
# When using Machine Learning

- Large, multi-dimensional and heterogeneous dataset
- Aimed at extracting knowledge
  - we know a pattern exists
- Complex task or problem for which there is not a (recommended) mathematical approach
- Large collections of data are available
  - data driven approach
- Limited prior knowledge

.pull-left[]
.pull.right[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='right', out.width="50%"}
library(knitr)
img_path <- "images/When_ML.png"
include_graphics(img_path)
```
]
---
class: center, middle, inverse

# Some examples

---

# Predicting cardiovascular disease with EHR 

.pull-left[

- 681 UK General Practicer
- 383,592 patients free from CVD (since 1st Jan 2005)
- Two-fold cross validation (similar to other epidemiological studies)
    - n = 295,267 _training set_
    - n = 82,989 _validation set_
- 30 separate included features
    - including biometrics, clinical history, lifestyle, test results, prescribing
- Four types of models: logistic, random forest, gradient boosting machines, and neural networks
]



.pull-right[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width="90%"}
library(knitr)
img_path <- "images/ML-RiskClinicalData.png"
include_graphics(img_path)
```

]

---
# Extending searches for SRs with SVM

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width="90%"}
library(knitr)
img_path <- "images/JCE.png"
include_graphics(img_path)
```

---
# Extending searches for SRs with SVM

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width="100%"}
library(knitr)
img_path <- "images/JCE_table2.png" 
include_graphics(img_path)
```

---
# ML for predicting readmission to pediatric ED

.pull-left[
- 28,341 children aged 0-14 accessing the ED of Regina Margherita Children's Hospital 
  - between October 2012 - April 2013
- EHR recorded: 
  - age, sex, nationality 
  - day of the week, of hospital admission, time of arrival, triage color
  - ICD9, final diagnosis, medication prescribed and diagnostic test
  - mode of discharge (at home, to a ward, ...), time lapse between 1st and 2nd visit
- __Aim:__ predicting risk of early readmission to ED (within 120 hours)
]

.pull-right[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width="115%"}
library(knitr)
img_path <- "images/HospitalReadmission.png" 
include_graphics(img_path)
```
]


---
# ML for predicting readmission to pediatric ED
- ANN's Architecture

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width="80%"}
library(knitr)
img_path <- "images/ANN.png" 
include_graphics(img_path)
```

- Accuracy: 91.3%
- Specificity: 98.3%
- Sensitivity: very low
---
# ML for predicting readmission to pediatric ED

- Tree
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width="95%"}
library(knitr)
img_path <- "images/CT.png" 
include_graphics(img_path)
```
- ACC: 81%; SPEC: 97%; SEN: 79.8%

---
class: center, middle, inverse

# Potential applications to chronic disease conditions

---
# Q-learning in healthcare

-  Growing segment of elderly living longer with chronic health conditions 
    - more likely to experience multiple transitions from hospital to home that negatively affect their quality of life and consume substantial healthcare resources

- Analyses attempting to identify patients at risk of readmission are mainly focused on the single re-hospitalization 
    - they fail to incorporate multiple hospital readmissions and admissions in community services that characterize a chronic condition
       
- Need of multistate models that considers all clinical pathways and  assesses their potential dependence on covariates 
  - multiple hospitalization readmission, home/residential care activations 

---
# Aims

- Assessing the effect of certain clinical conditions in a community setting on multiple readmissions by including intermediate care services  
- Enhancing understanding of the clinical pattern of patients with chronic conditions for adverse prognosis 
    - it can help implementing appropriate intervention strategies 
    
- Identifying the global risk profile of patients with clinical complexities according to their transition patterns 
  - Identifying clinical factors associated with different transitions, i.e. admission, discharge and death 
- To date, multi-state (statistical) models have been applied for estimating both competing and recurrent events
  - oncological patients 

---

# Transitions and time-varying confounders

- Effect of disease condition (T0, T1) on the outcome variable Y

.right-column[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='right', out.width="90%"}
img_path <- "images/figure_2.bmp"
include_graphics(img_path)
```
]
---
# Transitions and time-varying confounders

- Effect of disease condition (T0, T1) on the outcome variable Y
- Condition in the first period (T0) affects the health status in the next period (L1)

.right-column[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='right', out.width="90%"}
img_path <- "images/figure_3.bmp"
include_graphics(img_path)
```
]
---
# Transitions and time-varying confounders

- Effect of disease condition (T0, T1) on the outcome variable Y
- Condition in the first period (T0) affects the health status in the next period (L1)
- Health status (L1) affects the condition in the next period (T1) $\ldots$


.right-column[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='right', out.width="90%"}
img_path <- "images/figure_4.bmp"
include_graphics(img_path)
```
]
---
# Transitions and time-varying confounders

- Effect of disease condition (T0, T1) on the outcome variable Y
- Condition in the first period (T0) affects the health status in the next period (L1)
- Health status (L1) affects the condition in the next period (T1) $\ldots$
- $\ldots$ and the outcome $Y$

.right-column[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='right', out.width="90%"}
img_path <- "images/figure_5.bmp"
include_graphics(img_path)
```
]

---
# Inverse Probability of Treatment Weighted 

.pull-left[
- Estimate the outcome-exposure mechanism
    - How outcome-exposure mechanism depends on the observed past?
    - Use it to reweight data
]
.pull-right[
- Estimate the _propensity score_ for each time
$$g_t = P(T_t\vert L_t, T_{t-1})$$
- Estimate the conditional probability of observed history
$$g_{0:t^*} = \Pi_{t=0}^{t^*}g_t$$
- Inverse weighting of observed outcome with $g_{0:t^*}$
]

---
# Targeted Machine Learning

.pull-left[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width="100%"}
library(knitr)
img_path <- "images/TMLE.png"
include_graphics(img_path)
```
]

.pull-right[

$$\Psi(P_0) = E_{X,0}[E_0(Y|T=1, X) - E_0(Y|T=0, X)]$$

$$\Psi(Q_n) = \frac{1}{n}\sum_{i=1}^n[Q^0_n(1, X_i)-Q^0_n(0, X_i)]$$
__Objective:__ learn the target parameter of $P_0$ denoted by $Q(P_0)$:
- loss function L such that $Q_0$ minimized the risk $P_0L(Q)\equiv E_0L(Q)(0)$
]

.footnote[Targeted Learning, MJ van der Laan, S. Rose 2011]

---
# Two stage methodology
1. __Super learning__ (M.J van der Lan et al. 2007, E. Polley et al 2012)
  - uses a library of candidate estimators (parametric models and machine learning algorithms, i.e. neural networks, RandomForest, etc)
  - builds data-adaptive weighted combination of estimators using cross validation (_ensemble model_)

2. __Targeted estimation__ (M.J van der Lan et al. 2006)
  - Update initial estimate to remove bias for the parameter of interest 
  - Update based on parametric submodel through initial estimate with score equal to efficient score/efficient influence curve
  - Calculate final parameter from updated fit of the data-generating distribution
---
# Targeted Maximum Likelihood Estimation
- Regress outcome on past
$$Y\sim L_0 + L_1 + T_0 + T_1$$
  - take predictions setting $T_0=t_0$ and $T_1=t_1$
  - update predictions using $g_{0:1}=\Pi_{t=0}^1 g_t$

- Regress updated prediction $\hat Y^2$ on past to previos period
    - $\hat Y^2 \sim L_0 + T_0$
    - take predictions setting up $T_0$
     - update predictions using $g_0$
- Take average of prediction $\hat Y^1$ over $L_0$

---
# Properties

.pull-left[
- It requires you specify the correct propensity score model
  - If outcome-exposure mechanism estimated using a misspecified model
- Subject to bias and high variance with moderate to strong confounding
  - When certain treatment or exposure levels of interest are rare/absent for some patients
]
.pull-right[
- Minimize bias due to model misspecification
  - Double robust
  - Consistent 
- Maximize precision of effect estimates
  - Efficient (minimal asymptotic variance in semi-parametric model)
- May reduce bias and variance compared to IPTW
- Exploits Machine Learning for model specification
]


---
class: center, middle

# Thanks!

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

The chakra comes from [remark.js](https://remarkjs.com)






